---
title: "Lab 05 - La Quinta is Spanish for next to Denny's, Pt. 2"
author: "Benjamin Egan"
date: "02-12-25"
output: github_document
---



```{r load-packages, message = FALSE, warning = FALSE, echo = FALSE}
library(ggrepel)
library(tidyverse) 
library(dsbox) 
```

```{r load-data, message = FALSE, echo = FALSE}
states <- read_csv("data/states.csv")
```

### Exercise 1

```{r saving headache on backend, echo = FALSE}
dennys <- dennys %>%
  mutate(establishment = "Denny's")
laquinta <- laquinta %>%
  mutate(establishment = "La Quinta")

```


```{r Alaska locations}
dn_ak <- dennys %>%
  filter(state == "AK")
nrow(dn_ak)

lq_ak <- laquinta %>%
  filter(state == "AK")
nrow(lq_ak)


```

There are 3 Denny's and 2 La Quintas in Alaska.

### Exercise 2

```{r how many combinations}
nrow(lq_ak) *nrow(dn_ak)
```

Since there are 3 Denny's and 2 La Quintas in Alaska, we need 6 total distances. We can combine the two data frames to aid in the calcuations.

```{r setup full data frame, warning = FALSE}


dn_lq_ak <- full_join(dn_ak, lq_ak, 
                      by = "state")
view(dn_lq_ak)
```

Once we combine the data frames, the dn_lq_ak data frame contains 6 rows, one row per combination. Each variable has an .x or .y designation to differentiate between if it is for a Denny's or a La Quinta. We can use mutate() to add a variable for distance.


### Exercise 3

```{r haversine calcuation, echo = FALSE}

haversine <- function(long1, lat1, long2, lat2, round = 3) {
  long1 <- long1 * pi / 180
  lat1 <- lat1 * pi / 180
  long2 <- long2 * pi / 180
  lat2 <- lat2 * pi / 180

  R <- 6371

  a <- sin((lat2 - lat1) / 2)^2 + cos(lat1) * cos(lat2) * sin((long2 - long1) / 2)^2
  d <- R * 2 * asin(sqrt(a))

  return(round(d, round))
}

```

```{r calcuate distance}

dn_lq_ak <- 
  dn_lq_ak %>%
  mutate(
    distance = haversine(longitude.x, latitude.x, longitude.y, latitude.y, 2)
  )

view(dn_lq_ak)

```



### Exercise 4

```{r minimum distance}

dn_lq_ak_mindist <- dn_lq_ak %>%
  group_by(address.x) %>%
  summarize(closest = min(distance))

dn_lq_ak_mindist <- dn_lq_ak_mindist %>%
  rename(address = address.x)

view(dn_lq_ak_mindist)
```

```{r setup for other states, warning = FALSE, echo = FALSE}

# North Carolina
dn_nc <- dennys %>%
  filter(state == "NC")

lq_nc <- laquinta %>%
  filter(state == "NC")

dn_lq_nc <- full_join(dn_nc, lq_nc, 
                      by = "state")
dn_lq_nc <- 
  dn_lq_nc %>%
  mutate(
    distance = haversine(longitude.x, latitude.x, longitude.y, latitude.y, 2)
  )

dn_lq_nc_mindist <- dn_lq_nc %>%
  group_by(address.x) %>%
  summarize(closest = min(distance))

dn_lq_nc_mindist <- dn_lq_nc_mindist %>%
  rename(address = address.x)


# Texas
dn_tx <- dennys %>%
  filter(state == "TX")

lq_tx <- laquinta %>%
  filter(state == "TX")

dn_lq_tx <- full_join(dn_tx, lq_tx, 
                      by = "state")
dn_lq_tx <- 
  dn_lq_tx %>%
  mutate(
    distance = haversine(longitude.x, latitude.x, longitude.y, latitude.y, 2)
  )

dn_lq_tx_mindist <- dn_lq_tx %>%
  group_by(address.x) %>%
  summarize(closest = min(distance))

dn_lq_tx_mindist <- dn_lq_tx_mindist %>%
  rename(address = address.x)

dn_lq_tx_mindist <- dn_lq_tx_mindist %>%
 mutate(
    close = if_else(closest <= 5, "Less than 5", "More than 5")
  )

# Georgia
dn_ga <- dennys %>%
  filter(state == "GA")

lq_ga <- laquinta %>%
  filter(state == "GA")

dn_lq_ga <- full_join(dn_ga, lq_ga, 
                      by = "state")
dn_lq_ga <- 
  dn_lq_ga %>%
  mutate(
    distance = haversine(longitude.x, latitude.x, longitude.y, latitude.y, 2)
  )

dn_lq_ga_mindist <- dn_lq_ga %>%
  group_by(address.x) %>%
  summarize(closest = min(distance))

dn_lq_ga_mindist <- dn_lq_ga_mindist %>%
  rename(address = address.x)
```



## Graphing The Distance

### Alaska
 
```{r comparison for Alaska}     

dn_lq_ak_mindist %>%
  ggplot(mapping = aes(
    x = fct_reorder(address, closest),
   y = closest
  )) +
  theme_bw()+
  
    geom_col(fill = "#6a4e93") +
  labs(
    title = "Minimum distance from a Denny's to a La Quinta in Alaska",
    x = "Address of a Denny's",
    y = "Distance to a La Quinta in km"
  )

mean(dn_lq_ak_mindist$closest)


```

This graph shows us that the closest La Quinta to a Denny's is only about 2 km away. The farthest La Quinta is only about 6 km away. I also calculated the average distance of a Denny's to a La Quinta to be 4.41 km.


### North Carolina

```{r comparison for North Carolina}     
title_nc <- "Minimum distance from a Denny's to a La Quinta in North Carolina"

dn_lq_nc_mindist %>%
  ggplot(mapping = aes(
    x = fct_rev(fct_reorder(address, closest)),
   y = closest
  )) +
  theme_bw()+
  coord_flip()+
  geom_text(aes(x = address, 
              y = closest, 
              label = closest),
          size = 3.7, color = "black") +
    geom_col(alpha = .4, fill = "#6a4e93") +
  labs(
    x = "Address of a Denny's",
    y = "Distance to a La Quinta in km"
  )+
  ggtitle(title_nc)+
  theme(plot.title = element_text(size=12.5))

mean(dn_lq_nc_mindist$closest)

```

This graph shows us that the same trend does not hold in North Carolina. The average distance from a Denny's to a La Quinta is 65.44 km, with only four locations under 10 km.


### Texas


```{r comparison for Texas, warning = FALSE}     
title_tx <- "Minimum distance from a Denny's to a La Quinta in Texas"
dn_lq_tx_mindist %>%
  ggplot(mapping = aes(
   x = closest
  )) +
  theme_bw()+
    geom_histogram(binwidth = 1,fill = "#6a4e93") +
  labs(
    x = "Distance to a La Quinta in km",
    y = NULL
  )+
  ggtitle(title_tx)+
  theme(plot.title = element_text(size=14))

mean(dn_lq_tx_mindist$closest)
sd(dn_lq_tx_mindist$closest)
dn_lq_tx_mindist%>%
count(close)

```

The average distance from a Denny's to a La Quinta is around 5.7 km, with the standard deviation being 8.83 km. By creating a variable that distinguishes distance, we can see that 126 La Quinta's are less than 5 km away from a Denny's, while 74 of them are over 5 km away.


### Georgia

Since I am from Atlanta, I decided to pick Georgia.

```{r comparison for Georgia, warning = FALSE}     
title_ga <- "Minimum distance from a Denny's to a La Quinta in Georgia"

dn_lq_ga_mindist %>%
  ggplot(mapping = aes(
    x = fct_rev(fct_reorder(address, closest)),
   y = closest
  )) +
  theme_bw()+
  coord_flip()+
  geom_text(aes(x = address, 
              y = closest, 
              label = closest),
          size = 3.7, color = "black") +
    geom_col(alpha = .4, fill = "#6a4e93") +
  labs(
    x = "Address of a Denny's",
    y = "Distance to a La Quinta in km"
  )+
  ggtitle(title_ga)+
  theme(plot.title = element_text(size=12.5))

mean(dn_lq_ga_mindist$closest)
sd(dn_lq_ga_mindist$closest)
```

In Georgia, mean is 14.05 km and the standard deviation is 20.62 km. You can see that the majority of Denny's have a La Quinta less than 20 km away.


### Final Question

I'm guessing this holds up the best in Texas. I didn't look at other states with high numbers of Denny's/La Quintas (i.e. California), but Texas has a ton of low distances based on the datasets.



## Visual by geographic location

I also decided to graph these relationships geographically, using longitude and latitude.

### Alaska
```{r creating dataset for Alaska visualization, echo = FALSE}

all_together <- rbind(dn_ak,lq_ak)

AK_dataset <- full_join(all_together, dn_lq_ak_mindist,
          by = "address")
```

```{r geographic for Alaska}     

AK_dataset %>%
  ggplot( mapping = aes(
    x = longitude,
    y = latitude,
    color = establishment
  )) +
  theme_bw()+
    geom_point() +
 scale_color_manual(values=c('Blue','Red'))+
geom_text_repel(aes(x = longitude, 
              y = latitude, 
              label = closest),
          size = 3, color = "black") +
  labs(
    title = "Minimum distance from a Denny's to a La Quinta in Alaska",
    x = "longitude",
    y = "latitude"
  )
```

This map is a geographical representation of the distance between Denny's and La Quinta. You can tell that they are tightly clustered together, indicating Denny's are always close to La Quintas.


### North Carolina

```{r creating dataset for NC visualization, echo = FALSE}

all_together <- rbind(dn_nc,lq_nc)

NC_dataset <- full_join(all_together, dn_lq_nc_mindist,
          by = "address")
```

```{r geographic for North Carolina, warning = FALSE}     

NC_dataset %>%
  ggplot( mapping = aes(
    x = longitude,
    y = latitude,
    color = establishment
  )) +
  theme_bw()+
    geom_point(alpha = .6) +
 scale_color_manual(values=c('Blue','Red'))+
geom_text_repel(aes(x = longitude, 
              y = latitude, 
              label = closest),
          size = 3, color = "black") +
  labs(
    title = "Minimum distance from a Denny's to a La Quinta in North Carolina",
    x = "Longitude",
    y = "Latitude"
  )
```

This graph is more messy and less helpful than the Alaska map. The increase in locations make it harder to understand which La Quinta each Denny's is going to, but it does have advantages. One primary advantage is understanding why there are a bunch of triple digit distances. You can see that the bottom right corner of the state has several Denny's without a La Quinta.


### Texas

```{r creating dataset for TX visualization, echo = FALSE}

all_together <- rbind(dn_tx,lq_tx)

TX_dataset <- full_join(all_together, dn_lq_tx_mindist,
          by = "address")
```

```{r geographic for Texas, warning = FALSE}     

TX_dataset %>%
  ggplot( mapping = aes(
    x = longitude,
    y = latitude,
    color = establishment
  )) +
  theme_bw()+
    geom_point(alpha = .6) +
 scale_color_manual(values=c('Blue','Red'))+
geom_text_repel(aes(x = longitude, 
              y = latitude, 
              label = closest),
          size = 3, color = "black") +
  labs(
    title = "Minimum distance from a Denny's to a La Quinta in Texas",
    x = "Longitude",
    y = "Latitude"
  )
```

This is a very messy plot and something I would absolutely not use. I am concerned about the datapoints on the far left of the graph not showing distances.

### Georgia

```{r creating dataset for GA visualization, echo = FALSE}

all_together <- rbind(dn_ga,lq_ga)

GA_dataset <- full_join(all_together, dn_lq_ga_mindist,
          by = "address")
```

```{r geographic for Georgia, warning = FALSE}     

GA_dataset %>%
  ggplot( mapping = aes(
    x = longitude,
    y = latitude,
    color = establishment
  )) +
  theme_bw()+
    geom_point(alpha = .6) +
 scale_color_manual(values=c('Blue','Red'))+
geom_text_repel(aes(x = longitude, 
              y = latitude, 
              label = closest),
          size = 3, color = "black") +
  labs(
    title = "Minimum distance from a Denny's to a La Quinta in Georgia",
    x = "Longitude",
    y = "Latitude"
  )
```

I'll admit this looks nothing like Georgia, and some of the data points written don't appear to be as far as they seem. But it's a cool visual.